---
title: "Discrete Data Analysis for Machine Learning Practitioners, with Examples in R"
author: "Julian Hatwell"
date: "7 February 2019"
output: html_document
references:
- id: noack1950
  title: A class of random variables with discrete distributions.
  author:
  - family: Noack
    given: A.
  container-title: Annals of Mathematical Statistics
  volume: 21
  page: 127-132
  type: article-journal
  issued:
    year: 1950
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE
                      , message = FALSE
                      )

knitr::opts_template$set(
  fig.wide = list(fig.height = 4.5, fig.width = 8, fig.align='center')
  , fig.wideX = list(fig.height = 3, fig.width = 9, fig.align='center')
  , fig.relaxed = list(fig.height = 6, fig.width = 8, fig.align='center')
  , fig.tile = list(fig.height = 3, fig.width = 3, fig.align='center')
)
```

# Introduction

The focus for today will be on visual analytics, not theories and formulas!

Note: This tutorial is based on material contained in the book [Discrete Data Analysis with R: Visualization and Modeling Techniques for Categorical and Count Data](https://www.crcpress.com/Discrete-Data-Analysis-with-R-Visualization-and-Modeling-Techniques-for/Friendly-Meyer/p/book/9781498725835) by Michael Friendly. Michael Friendly is a pioneer in this field and has contributed to the development of modules and libraries for SAS and R.

You can also find a much shorter tutorial on these topics in the long form help for the "vcd" package for R: [vcd vignette](https://cran.r-project.org/web/packages/vcdExtra/vignettes/vcd-tutorial.pdf)

## Intended Audience

Data Analytics and Artificial Intelligence Research Group at Birmingham City University. A team of academics, post-doctoral researchers and doctoral candidates.

## Why Would You Be Interested in This Workshop?

**Aims:** 

To provide a high-level overview of some less well-known, visual statistical tools that can be applied to very common tasks in Machine Learning practice and research.

**Objectives:**

* Present some visual, statistical tools that are not usually included in a formal stats curriculum
* Discuss how these could be applied to tasks that are typical in many practical Machine Learning projects
* Demonstrate this with relevant case studies:
    * Text/Event Analysis
    * Comparing True and Predicted Classes for Ordinal data
    * Exploratory Data Analysis of Categorical Data
    * Dimension Reduction and Clustering of Categorical Data

**Learning Outcomes:** After this session you will be able to:

* Demonstrate rigorous and robust, yet visually intuitive reporting of machine learning research results in:
    * Frequency data: events, counts, occurrences, wait-times
    * Multi-class problems involving ordinal data
* Formulate an appropriate strategy for the Exploratory Data Analysis (EDA) phase of a machine learning project:
    * Assess a new dataset that contains several discrete features
    * Develop an intuition about the relationships between those features

EDA is an essential but often overlooked step prior to any analysis task.

## What we won't cover

Bear in mind, this is just an overview! Our seminar session is just long enough for a taster and not much depth.

In this workshop, we will focus on intuitive, visual tools that can reveal patterns in categorical data. We will only introduce formal statistics and tests where necessary, and will not develop any mathematical notation, theory or proofs.

We also won't cover explanatory and predictive modeling. Most of you will already be familiar with logistic regression for binary classification. There are related or equivalent techniques for other kinds of classifiation (ordinal, polytomous/multi-class and counts). Even though these topics fall squarely under both DDA and Machine Learning, this is all out of scope for today because it's a massive topic that requires a whole course/module to explain.

## Why Am I Presenting This Workshop?

This branch of stats peaked my interest at two distinct points in my life:

1. My undergraduate degree in Microbiology - using techniques such as [Most Probable Number](https://en.wikipedia.org/wiki/Most_probable_number) to estimate the number of bacterial cells in a test tube. You certainly can't count them all!

2. My career as a data professional in the tertiary education sector.

The kinds of questions I used to work on were:

* Does the number of students who progress from undergrad to post-grad differ, depending on whether they receive an incentive plan?
* Do different interventions improve the retention of at-risk students?
* What combination of factors most affect student satisfaction (measured on a likert scale)
* How do student outcomes (Distinction, Merit, Pass, Fail) compare across various demographics?
* Can we identify groups in need of extra support?
    
Here are just some of the factors/features to be utilised in the research questions above:

* subject
* full-time / part-time
* age group
* gender
* nationality (well over 50 unique categories at a single institution)
* high school attended prior to entry
* subjects and grades prior to entry
    
The types of questions we did *not* need to ask was "what is the average height/weight of students across different groups?" So it was obvious that regular descriptive and quantitative stats (i.e. the normal distribution) were not the right tools for this work. 

Tables and cross tabs were no use with so many dimensions. Pivot tables were a little better, but couldn't help with signficance testing and predictive analytics. Most importantly, I needed visual tools to communicate findings with the senior management team. I couldn't expect those guys to squint over a page of tables and numbers with me, looking for clues. This is when I came upon Michael Friendly's work on visualising categorical data, and was highly motivated to learn more about it.

That's a story about the BI, management reports and forecasting work that I used to do. Now I am involved in (or evolved into) Machine Learning research and practice, and I face completely new tasks. However, I keep noticing that the old tools and techniques are still applicable. I wanted to share these ideas with you and I hope you also find them useful.

# Preliminaries

If you prefer, you can simply follow along with the in person presentation but if you want to run the code chunks and see the results in the RStudio UI you need access to a machine with R+RStudio and the source code for [this file](https://raw.githubusercontent.com/julianhatwell/DDAR/master/DDAR_Display_Page.Rmd). After you've downloaded it to your working directory, just ditch all my text move all the code chunks to a new .R file with:

```{r purlit, eval=FALSE}
knitr::purl("DDAR_Display_Page.Rmd")
```

To save time, make sure the following libraries are installed and loaded before the workshop.

```{r load_libs}
library(knitr)
library(vcd)
library(vcdExtra)
library(ca)
```

## What is Discrete Data Analysis (DDA)?

Predictably, the term refers to analysis of discrete data! DDA is a set of tools and techniques that you would apply to achieve the same objectives as any quantitative statistical analysis but with these data types:

* Nominal
* Ordinal
* Counts / Frequencies

Nearly everyone with any background in stats has some experience of DDA! We have all performed a $\chi^2$ test for independence or for association. This is a hypothesis test involving categorical and count data. For example:

* $H_0 =$ There is an equal number of instances belonging to each category of feature A. Any observed differences are small enough to be considered a sampling error.

* $H_0 =$ Membership of any category in feature A is independent of membership of any category in feature B. Frequency in each combination is proportional to the joint distribution $P(A) \times P(B)$. Any observed differences are small enough to be considered a sampling error.

## Common distributions - the currency of DDA

There are several distributions to help with the most common DDA research questions. You can recognise them because a random variable from any of these can only be a natural number $\{ 0,1,2,\dots \}$:

* Binomial - how many heads in n coin tosses?
* Poisson - how many buses come every n minutes?
* Geometric - how many minutes do I wait before a bus comes?
* Negative Binomial - how many minutes waiting time between each bus?

These distributions are all special cases of the Power Series family of distributions [@noack1950]

There are lots of other more specialised distributions such as *hurdles* and *zero-inflated* models where a very high prevalence zero success events must be handled appropriately.

## Common DDA Tasks

This is a non-exhaustive list of common statistical tasks and questions that require specific DDA techniques.

1. 1-way analysis and goodness of fit
    * $H_0 =$ My data fits a common, known distribution $\delta$, so I can use mathematical/analytical properties of $\delta$ for parametric tests.
    * Does my data fit a known distribution and if so, what are the parameters?
    * Does one random sample come from the same distribution as another?
2. 2-way analysis
    * 2 $\times$ 2 tables are a special case where a standard $\chi^2$ test is not recommended. What tools should I use instead?
    * $\chi^2$ tests on $n \times m$ tables can tell me that an association between features is present/absent, but if $n$ or $m$ are larger than 3-4, how can I find and test for complex relationships between so many combinations of categories?
    * Some of my categories are ordinal - what is the right test?
    * Classification problems with ordinal data - how can I quantify the agreement/disagreement between True and Predicted classes when there is a relationship between classes?
3. n-way analysis:
    * Visualising multi-dimensional tables for exploratory analysis
    * Statistical hypothesis testing in n-way tables
    * Correspondence Analysis, a dimension reduction technique, like PCA and clustering for categorical variables!
4. Explanatory and Predictive modeling:
    * Any of the above, measuring the magnitude of interactions
    * Any of the above, measuring the effect of covariates and co-factors

## DDA in R and task specific visualisations

We don't have time develop a fluency for these tasks in this workshop but if this is something useful for you there are some "building block" skills that are worth spending time on:

1. The most useful data structures for DDA in R are not data frames, but n-dimensinal arrays, n-way tables, and matrices, sometimes data frames with a frequency column (for the count data).

1. It's good to become fluent with functions for manipulating table data - pivoting and collapsing dimensions, and converting between data structures to suit your analysis. Useful functions to get to know are table(), addmargins(), prop.table(), ftable(), xtabs(), xtable(), structable(), apply(), margin.table(), collapse.table(), aggregate(), aperm(), expand.dft(), subset(), using integer indexing to re-order categories and pivot dimensions. Most of the specialised objects have as.table() and as.data.frame() methods that generate the most intuitive shape for the output data structure.

1. Library "vcd" comes with its own plotting system. It is based on the grids package but is different from R base graphics, ggplot2 and lattice! It's easy to get started, but a bit of effort is required to learn all the tweaks for making publication ready graphical output. However, interpreting these area-based plots is really intuitive and it's worth the effort to learn so that results can be communicated succinctly (a picture tells a thousand words).

# Workshop

## Task 1: 1-way analysis (goodness of fit) for monitoring the frequency of events.

Count, or frequency data has many varied applications from event monitoring to text analysis. 1-way analysis uses distributions to compare different samples, and monitor for changes in event frequency.

For the first task we will look at two common distributions for count data and compare two samples.

The poisson distribution has just one paramater $\lambda$ for the rate, and both the mean and variance are equal to this parameter.

The negative binomial distribution has two parameters, rate and dispersion. The mean is equal to the rate, but the variance is always greater than the rate. It can appear as a right skewed poisson distribution and behaves as if a poisson distribution is being sampled over a period of time while the rate is non-constant.

However, data from these two distributions can look very similar. Identifying the best match and estimating the parameters is essential for accurate monitoring and reporting results.

### Case Study in Text Analysis: Marker word frequency

The following technique has been used to determine if documents are written by same or different authors. Distributions of marker words are fit to text that is known to come from one author. 

Any previously unseen text can then be tested against these known distributions to determine if they are written by the same author.

In this simple case study, we will use the Federalist dataset to analyse how many times the word 'may' appear per 200 word block. This is a marker word, that could help identify an author by how often they use it. Most studies of this kind would compare results across many such tests with different words.

Notes:

* The rate is occurences/200 words. 
* The expected count is the rate parameter $\times$ 200
* Plotting the square root transformation of the frequency provides more detail on the small numbers in the right tail of these charts. Log transformation would also work but the rootogram visualisations use square root transformation by default. So, this is also applied manually where necessary for comparison.

```{r federalist}
data("Federalist", package = "vcd")
Federalist
sum(Federalist) # number of blocks
sum(expand.dft(as.data.frame(Federalist))$nMay) # number of occurences
mean(expand.dft(as.data.frame(Federalist))$nMay)
var(expand.dft(as.data.frame(Federalist))$nMay)
```

The discriptive stats showed a slightly higher variance but we don't know if this is significant. What's the right test to use here?

A very naive approach is to take the mean and visually compare a poisson distribution with this value for $\lambda$ against the actual data.

```{r pois_naive_expected}
barplot(sqrt(200 * dpois(0:6, mean(expand.dft(as.data.frame(Federalist))$nMay)
))
        , main = "Number of text blocks by Number of Occurrences\nnaive expected values"
        , ylab = "sqrt(Naive Expected)"
        , xlab = "Occurences of 'may'"
        , col = "lightblue")
```

The actual data looks a bit like this distribution. Could this be a good fit? How can we easily compare them?

```{r barplot_actual}
barplot(sqrt(Federalist)
        , main = "Number of text blocks by Number of Occurrences\nactual values"
        , xlab = "Occurences of 'may'"
        , ylab = "Sqrt(Actual)"
        , col = "lightgreen"
)
```

Use the goodfit function in the vcd package to estimate the parameters and compare the fit of different distributions. It provides hanging rootograms that are intuitive views of discrete goodness of fit testing.

First, a poisson fit using maximum likelihood - exactly what we've just done, but with bells on it:

```{r fed_goodfit_pois_chisq}
Fed_fit0 <- goodfit(Federalist, type = "poisson")
# This will show the rate parameter, estimated by Maximum Likelihood
# This estimate is the same as the naive mean. This is to be expected when fitting a poisson.
unlist(Fed_fit0$par)

# This will show a Chi-Square Goodness of fit test against the expected values of a poisson distribution with this parameter
summary(Fed_fit0)
```

The $\chi^2$ test is very strong evidence to reject the null hypothesis. This sample is not from a poisson distribution with rate = sample mean. However, we don't know much about what caused this result. However, the rootogram plot makes it very clear what the problem is by showing how the variable diverges from the expectations. 

```{r fed_goodfit_pois_plot}
plot(Fed_fit0, type = "hanging", shade = TRUE)
```

By hanging each bar top down from the expected distribution, we just need to look at how the bottoms of the bars miss the $y = 0$ reference line. This is visually much more intuitive than comparing the tops of the bars to a curve. The contribution of each bar to the $\chi^2$ statistic is also made clear by the shading.

We can try fitting a poisson using a different estimation method, minimising the $\chi^2$ errors instead:

```{r fed_goodfit_pois_mchisq}
Fed_fit0 <- goodfit(Federalist, type = "poisson", method = "MinChisq")
# This has found a different value for the rate parameter.
unlist(Fed_fit0$par)

# This will show a Chi-Square Goodness of fit test against the expected values of a poisson distribution with this parameter
summary(Fed_fit0)
```

The $\chi^2$ test is also very strong evidence against the null.

```{r fed_goodfit_pois_mplot}
plot(Fed_fit0, type = "hanging", shade = TRUE)
```

The rootogram shows that the residuals have just been shuffled around so they balance better by sign. Clearly poisson is not the right choice.

Next, fit a negative binomial fit:

```{r fed_goodfit_nbinom}
Fed_fit1 <- goodfit(Federalist, type = "nbinomial")
# This will show the rate and dispersion parameters, estimated by Maximum Likelihood
# prob is the rate. Here it's very close to the poisson mean
# size is the dispersion
unlist(Fed_fit1$par)
summary(Fed_fit1)

plot(Fed_fit1, type = "hanging", shade = TRUE)
```

We have estimated some parameters and the $\chi^2$ goodness of fit test result is not significant. We accept that null hypothesis that this sample is from a negative binomial distribution having the estimated parameters. The plot speaks for itself now that the bars drop close to the $y = 0$ reference line, with only insignificant residuals.

We will now assess a different body of text.

```{r nonFederalist}
# Some dummy data
nonFederalist <- structure(c(135L, 71L, 36L, 17L, 8L, 4L, 1L), .Dim = 7L, .Dimnames = structure(list(
    c("0", "1", "2", "3", "4", "5", "6")), .Names = "nMay"), class = "table")
nonFederalist
sum(nonFederalist) # number of blocks
sum(expand.dft(as.data.frame(nonFederalist))$nMay) # number of occurences
mean(expand.dft(as.data.frame(nonFederalist))$nMay)
var(expand.dft(as.data.frame(nonFederalist))$nMay)
barplot(sqrt(nonFederalist)
        , main = "Number of text blocks by Number of Occurrences\nunseen text"
        , xlab = "Occurences of 'may'"
        , ylab = "Sqrt(Actual)"
        , col = "lightgreen"
)
```

The descriptive statistics are a bit larger than Federalist, but is this significant? We can test it properly using the parameters from the goodfit object we just created:

```{r fit_nonfed_chisq}
# we pass in our fitted parameter list this time
Fed_fit3 <- goodfit(nonFederalist, type = "nbinomial", par = Fed_fit1$par)
summary(Fed_fit3)
```

The $\chi^2$ result is significant, and strong evidence that this new text is not from the same distribution. However, on it's own, the test can't provide detailed information.

```{r fit_nonfed_plot}
plot(Fed_fit3, type = "hanging", shade = TRUE)
```

The rootogram hangs the frequency bars off the expected distribution, which this time is fixed with parameters we estimated on the original text. We can now see from the pattern of residuals that the word 'may' seems to appear more frequently in this text. Specifically, the tail, with 3-5 occurences per 200 words is fatter, while blocks where this word is absent are fewer.

### Talking points

* Does this test imply that the second text was not written by the same author?
* Text analysis has come a long way since these statistical techniques were developed. What are some newer, more sophisticated methods?
* Can you describe, in plain English, what is the implication of finding a poisson distributed word frequency? Why is the negative binomial a more natural distribution for word frequency? Hints:
    * Frequency is the number of times the word appeared per block of 200 words
    * Review the notes about these two distributions at the top of this task

## Task 2: 2-way analysis for reporting results for multi-class classification of ordinal variables.

In Machine Learning, the standard way to assess the quality of a supervised classification model is with a confusion matrix. This is a two way table comparing True and Predicted class labels. When classes are well balanced, it's easy to measure accuracy as $\frac{\sum(\textit{diagonal})}{\sum(\textit{total})}$. When they're out of balance you can have a situation such as this:

The ratio of class 1 : class 2 is 9:1. Predict every instance as class 1 and you will get 90% accuracy. This not a useful result!

So we use Cohen's $\kappa$ statistic to weight according to the occurence of each class in the data. $\kappa = 1$ means perfect agreement while $\kappa = 0$ means no better than chance agreement. The above example would score $\kappa = 0$. 

That's all well and good, but what if class labels are related to each other? If they have a natural order such as [*least, middle, most*] then you have an additional complication. You want your model to predict as well as possible, but to make an off-by-one error isn't as bad as a full opposite error. Also, the classification distribution might be biased, giving different skew and kurtosis (flattened or polarised) by the predictive model.

If you can detect this kind of bias, you can correct it! This is where agreement plots can help. They are based on statistical techniques for comparing two raters, such as two doctors, two film critics, two wine critics etc, to valdiate ratings and analyse these kinds of bias.

### Case Study: Comparing inter-rater agreement.

In the following example we show how two sets of diagnoses for Multiple Sclerosis compare between two doctors, one in Winnipeg and one in New Orleans. The classes are ordinal ratings:

1. Certain
1. Probable
1. Possible
1. Doubtful

```{r MSP_2D}
# combine results from two cities (same raters are involved in all cases)
MSP <- margin.table(MSPatients, 1:2)
```

On this dataset, we can't say which is the true and which is predicted because they're both independent raters, but for now let's assume that Winnipeg Neurologist is the true class and New Orleans Neurologist is the predicted class. The highest accuracy we could expect to get with constant model that always makes the same rating is the proportion of the majority true class. This would give a score of  $\kappa = 0$:

```{r MSP_winntrue}
margin.table(MSP, 2)
props <- prop.table(margin.table(MSP, 2))
props

# The baseline accuracy is
props[which(props == max(props))]
```

Without looking at the confusion matrix, let's take a naive look at the accuracy:

```{r MSP_accuracy}
# basic accuracy
sum(diag(MSP))/sum(MSP)
```

On the face of it, this looks like a terrible result! Is New Orleans Neurologist just predicting the most frequent Winnipeg class?

```{r MSP_confmat}
MSP
```

There are definitely some problems with the inter-rater agreement but it doesn't look like random guessing. Why is accuracy so low? Does class imbalance play a part? What is the $\kappa$ score?

```{r MSP_kappa}
# cohen's kappa - the function in the vcd package provides weights and confidence intervals for this statistic
confint(Kappa(MSP, weights = "Fleiss-Cohen"))
```

Unweighted $\kappa$, the standard measure, also shows that this is not a very good result. 

However, Fleiss-Cohen Weighted $\kappa$ favours near disagreements and taking this into account gives a much better result. Off-by-n errors appear to be common, so these raters are closer than it first appears.

An agreement plot is the best way to visualise a situation like this because it shows how off-by-n errors and other bias could be affecting the outcome.

```{r MSP_agreement, results='hold'}
op <- par(mar = c(4, 3, 4, 1) + .1)
B <- agreementplot(MSP
                   , main = "MS Patient Ratings"
                   , xlab_rot = -20
)
```

```{r, echo=FALSE}
# ignore this
par(op)
```

This plot needs some explaining. The rectangle outlines show the maximum possible agreement, given the marginal totals. The black rectangles indicate where there is full agreement between raters. The shaded area indicates the off by $n$ disagreement. *Note, if there were more categories, there would be more grades of shading*.

Ideally, we want the black area to fill the outline and it's not very close in this chart. However, the shaded area is coming much closer to filling the available space. 

```{r MSP_bangdi}
unlist(B)[1 : 2]
```

The Bangdiwala statistic is calculated based on the ratio of these shaded rectangles to the maximal outline. The large interval between the weighted and unweighted statistics shows that near disagreements are very common and if taken into account, they indicate a much better alignment than was indicated by a naive accuracy measure and even $\kappa$. 

There is another feature of this plot, which is the relationship of the rectangles to the diagonal line. This indicates how well aligned the raters are in using each category. In this chart, we can see how the Winnipeg Neurologist rates cases as "Certain" or "Probable" much more frequently than New Orleans Neurologist. Neither $\kappa$ nor Bangdiwala statistics can detect this. There are statistical tests available but this chart shows the problem directly.

With a little intervention, guided by this information, these two raters could become much better aligned.

### Talking Points

* What other statistics, tools and techniques can you use to analyse a confusion matrix?
* Have you worked with, or can you think of an ordinal classification problem?
* How could you use the Bangdiwala statistic *during* training to improve alignment?

## Task 3: n-way analysis for exploring relationships between features

Exploratory Analysis is an essential, and frequently overlooked step as practitioners are often keen to use the latest tooling or are under pressure to deliver results. However, spending some time "getting to know" a new dataset and developing some intiution about important features and interactions will pay off by revealing a better strategy when it comes to modelling.

Mosic plots are very powerful for visualising categorical data. Everything is based on the simple concept that your categorical features of interest are on the axes, and the area of plotted shapes is proportional to the count or frequency at the intersection of each category. Then, shading can be used in many ways to highlight important interactions. Today we will only use the mosaic plot, but this is the most useful as it scales well to 3 or even 4 dimensions.

*Note, the 'strucplot framework' that is the heart of the vcd package requires learning a new approach to structuring data and writing plotting routines compared to previous experience with R. However, it's worth the effort for the intuitive views of complex relationships in the data.*

### Case Study: Exploring the relationships between hair colour, eye colour and gender.

The HairEyeColor dataset ships with base R. It is a survey of 592 statistics students at the University of Delaware reported by Snee (1974). It is a 3-D array cross-tabulating the observations. The variables and their levels are as follows:

```{r describe_haireye, echo=FALSE}
hr_desc <- data.frame(No=1:3
                      , Name=c("Hair", "Eye", "Sex")
                      , Levels=c("Black, Brown, Red, Blond", "Brown, Blue, Hazel, Green", "Male, Female"))
kable(hr_desc)
```

For now, we will ignore the sex of each individual by collapsing the 3-D table down to 2-D, and re-order the table from dark to light eye colour (hair is already in that order).

```{r haireye}
haireye <- margin.table(HairEyeColor, 1:2)
haireye <- as.table(haireye[, c("Brown", "Hazel", "Green", "Blue")])
haireye
```

Big numbers and small numbers stand out but the table still has 16 cells which is a lot to take in at once. It's not immediately obvious what the relationships are.

Let's recall our old friend the $\chi^2$ test, which compares actual counts to expected counts. A non significant result would indicate that the two variables have no relationship (the null hypothesis).

```{r hr_chisq}
chisq.test(haireye)
```

The null hypothesis is rejected. There is a very significant difference between actual and expected counts. Obviously there is a relationship between these two features, but we still don't know what it is.

Expected counts is what we would get if there were no interaction between the features. Here's what no interaction looks like:

```{r hr_expected}
expected = independence_table(haireye)
round(expected, 1)

mosaic(expected
      , shade = TRUE
      , main="Expected frequencies"
      , labeling = labeling_values
      , value_type = "expected"
      , gp_text = gpar(fontface = 1))
```

Note that the areas are proportional to the counts. The grid lines between the tiles are dead straight because each cell is proportional to the marginal totals.

Here's what actuals looks like:

```{r hr_actuals}
mosaic(haireye
      , gp = shading_Friendly # shade = TRUE
      , main="Actual frequencies"
      , labeling = labeling_values
      , value_type = "observed"
      , gp_text = gpar(fontface = 1)
      , rot_labels = c(top = -20))
```

Here the grid lines between cells are no longer continuous because the cells are out of proportion with the marginal totals. Unexpectedly large cells show that the combination is over-represented and small cells the reverse.

The beauty of the mosaic chart is the use of shading to highlight the $\chi^2$ residuals. This shading scheme uses fills for statistically significant residuals but also adds an outline colour to highlight the sign of even the non-significant residuals. This is visually very powerful. There is a clear "opposite corner" pattern.

Let's take this up to all three dimensions now.

```{r hec}
HEC <- HairEyeColor[, c("Brown", "Hazel", "Green", "Blue"), ]
mosaic(HEC
      , gp = shading_Friendly # shade = TRUE
      , main="Actual frequencies"
      , labeling = labeling_values
      , value_type = "observed"
      , gp_text = gpar(fontface = 1), rot_labels = c(right = -45)) 
```

We can still see the general opposite corner pattern, and now some additional detail regarding gender. 

This type of plot is still fairly intuitive at 4 dimensions. More dimensions can be added by other means. There are several options:

* A cotabplot allows you to stratify over a further dimension. It looks just like facet_wrap() and facet_grid() in ggplot2, or a lattice/trellis plot.
* Mosaics can be featured in a plot matrix or pairs plot, where each feature is compared pairwise and inset on a grid. This is the closest thing to a splom in lattice, or a correlogram for continuous variables.
* Manipulate the grid spacing between dimensions to put more white space between several 4-way mosaics and a 5th dimensional slice. This is probably the most  fiddly option as you usually have to manage all the label placements, or suppress them.

There is a whole methodology around loglinear modeling which allows you to make statistical hypothesis tests about complex n-way relationships between features. There is enough material to write several books on this topic!

### Talking points

* Recall that we've made sure the categories are ordered light to dark. There is a very pronounced "opposite corner" relationship. What do you think that reveals (in plain English) and does this fit with intuition?
* Can you see a difference between males and females? If so, can you explain it? *Note, the original study was a self-reporting survey. The participants categorised themselves.*
* Why do the grid lines remain aligned horizontally between hair categories on both actual frequencies mosaics?
* If you discovered important relationships in categorical features during your exploratory phase, how would this affect your Machine Learning strategy?

## Task 4: Correspondence Analysis for dimension reduction and clustering

Most Machine Learning practitioners are familiar with dimension reduction. It can radically simplify and compress a dataset, making it easier to handle and train models on. Many dimension reduction techniques also double up as unsupervised learning methods i.e. clustering. This is useful as an end in itself, as well as being an additional tool for exploratory data analysis: multi-dimensional data is projected into 2-D which makes it intuitive to visualise.

If you already know something about Principle Components Analysis, this is similar, but not quite the same. Remember, in DDA, the "dimensions" we want to reduce or cluster are not only the features, but also the categories within each feature. PCA identifies new dimensions that capture the most variance. The equivalent concept in Correspondence Analysis is the "inertia" of the new dimensions.

Correspondence Analysis is an advanced topic. In particular, the way you pivot the multi-dimensional array controls the analysis. This requires some understanding of the data manipulation and modeling techniques. For today, we will just focus on interpreting the plots.

### Case Study: Analysis of Audience Viewing Data

The audience viewing data from Neilsen Media Research for the week starting November 6, 1995

It is a 3-D array cross-tabulating the viewing figures for three networks, between 8-11pm, Monday to Friday. The variables and their levels are as follows:

```{r describe_tv, echo=FALSE}
tv_desc <- data.frame(No=1:3
                      , Name=c("Day", "Time", "Network")
                      , Levels=c("Monday, Tuesday, Wednesday, Thursday, Friday", "8, 9, 10", "ABC, CBS, NBC"))
kable(tv_desc)
```

```{r tv_dataprep}
data("TV", package = "vcdExtra")
# The original data is collected in 15 minutes slices.

# Convert 3-D array to a frequency data frame
# This has a row for each cell of the array
# and a new column for the cell value
TV.df <- as.data.frame.table(TV)

# Convert it into hourly slices
levels(TV.df$Time) <- rep(c("8", "9", "10"), c(4,4,3))

# Convert frequency data back to 3-D array, now with just 3 time levels
TV3 <- xtabs(Freq~Day+Time+Network, TV.df)
```

```{r tv_3wayca}
# create a multiple correspondence analysis
TV3.mca <- mjca(TV3)

# the plot function uses all base R plot stuff
# but needs a bit of manipulation
cols <- c("blue", "black", "red")

# "blank plot"
res <- plot(TV3.mca, labels=0, pch='.', cex.lab=1.2)

# combine Dims, factor names and levels
coords <- data.frame(res$cols, TV3.mca$factors)

# hard-coded from known number of levels
# day, time, network
nlev <- c(5,3,3)

# everything needs to be in semantic order
coords <- coords[ order(coords[,"factor"], coords[,"level"]), ]
# quick fix for ordering e.g. day of week, not alphabetical
coords$order <- c(5, 1, 4, 2, 3, 6, 7, 8, 11, 9, 10)
coords <- coords[order(coords[, "order"]), ]

# place the points with separate plot chars and colours
points(coords[,1:2], pch=rep(16:18, nlev), col=rep(cols, nlev), cex=1.2)

# place the text
pos <- c(1,4,3)
text(coords[,1:2], labels=coords$level, col=rep(cols, nlev), pos=rep(pos,nlev), cex=1.1, xpd=TRUE)

# join things in sequence
lines(Dim2 ~ Dim1, data=coords, subset=factor=="Day", lty=1, lwd=1, col="blue")
lines(Dim2 ~ Dim1, data=coords, subset=factor=="Time",  lty=1, lwd=1, col="red")

# add segement from the origin to channels
nw <- subset(coords, factor=="Network")
segments(0, 0, nw[,"Dim1"], nw[, "Dim2"], col = "black", lwd = 0.5, lty = 3)

# add a legend
legend("topright", legend=c("Day", "Network", "Time"),
       title="Factor", title.col="black",
       col=cols, text.col=cols, pch=16:18,
       bg="gray95")
```

This plot reveals an incredibly strong effect relating NBC and Thursday nights. This is apparently most of the inertia captured by Dimension 1. What's going on there?

```{r TV_day_network}
margin.table(TV3, 1)
margin.table(TV3, c(1, 3))
```

Viewing figures are at their highest on Thursdays compared to any night of the week and for NBC they dramatically outnumber the other two channels, but only on this night. NBC has viewing figures that are comparable to the others most other nights. Everyone appears to be glued to NBC all evening. A little digging and domain knowledge (showing my age, here) means we can give this effect a name: This is the "Friends, Seinfeld and ER" effect! These were three of the most popular shows of that [era](https://en.wikipedia.org/wiki/1995%E2%80%9396_United_States_network_television_schedule). 

The 9pm slot clusters closer to the other four days than 8pm and 10pm. What's happening there?

```{r TV_time_network}
margin.table(TV3, 2)
t(TV3[4,,]) # Thursday
```

It turns out that 9pm is the most viewed timeslot throughout the week, independent of day or network. Yet, notice how the NBC audience numbers dip just a little for that time slot on Thursday, even though they still have far greater numbers than either of the other two channels. This dip in numbers runs counter to the "Friends, Seinfeld and ER" effect which has competely dominated the analysis, and so it has also been captured on Dimension 1.

Dimension 2 is forced to capture differences between ABC and CBS schedules on the other nights as well as possible, but there is undoubtedly more detail to untagle.

Now we have identified this difficulty, we can tackle the problem by making a joint analysis of time and day. This will produce a plot with a point for each time slot over the week and should tease apart the effect of people channel hopping from the "Friends, Seinfeld and ER" effect. It turns out to be very easy to do (see the code for yourself).

```{r tv_2wayca}
# Flatten to 2-D by stacking Time onto Network
# Note: The data shaping choice here controls the specifics of the analysis.
TV3s <- as.matrix(structable(Network~Time+Day, TV3))

# Create the Correspondence Analysis objects
TV3s.ca <- ca(TV3s)

# Generate the plot
res <- plot(TV3s.ca)
# add some segments from the origin to make things clearer
segments(0, 0, res$cols[,1], res$cols[,2], col = "red", lwd = 1)
segments(0, 0, res$rows[,1], res$rows[,2], col = "blue", lwd = 0.5, lty = 3)
```

As you can see, the small channel hop from NBC on Thursday has moved onto Dimension 2, along with the other variations in within-evening viewing patterns. On the previous plot, there is a very slight affinity 8pm-CBS, and 9pm-ABC. Did you notice? Look again. We can see clearly on this new plot, something of great interest at 8pm, Monday, but not later on the same evening and not again at the same time later in the week. We also can see several 9pm slots clustered around ABC.

An interesting feature of this new analysis, is that distance from the origin is proportional to the size of positive residual. What does this mean? Well for example, from 10pm the viewing figures dwindle, independent of day and channel as people go to bed or go out for a late drink. Thursday night, 10pm, NBC has massive audience figures relative to what's normally expected for that time slot the rest of the week. This inertia moves the point furthest from the origin out of all the nightly time slots. Monday at 8pm on CBS is notable for the same reason. Why not Monday 9pm and ABC? There are other nights when ABC is popular at 9pm, making this combination less unusual.

```{r TV_monday}
t(TV3[1,,]) # Monday
TV3[,2,1] # Every day, 9pm, ABC
```

As you can see Correspondence Analysis is an extremely powerful dimension reduction and clustering tool for categorical data.

# Summary

* We have demonstrated some classical statistical techniques that are not usually included in a basic stats curriculum.
* We have discussed how these techniques can be applied to tasks that are typical in many practical Machine Learning projects.
    * Text/Event Analysis
    * Comparing True and Predicted Classes for Ordinal data
    * Exploratory Data Analyisis
    * Dimension Reduction and Clustering
* We have shown that simple R functions and visual analyses are extremely powerful and you don't have to dive deep into stats to get the benefits.
* Nearly all the resources you need to take the next steps yourself are available in the "vcd" and "vcdExtra" package for R, R~/vignette("vcd")
* If you want to go much, much deeper with these topics, [Michael Friendly's book](https://www.crcpress.com/Discrete-Data-Analysis-with-R-Visualization-and-Modeling-Techniques-for/Friendly-Meyer/p/book/9781498725835) pulls together decades of research in this field and is a one-stop reference for this branch of statistics.

In our rush to use the latest developments and technologies in Machine Learning and Deep Learning, it's easy forget that none of it would be possible without pioneering statistical research of previous decades. Most of this work is still relevant to everyday problems in Machine Learning practice, and it's always worth a review of the foundations before reaching for the newest tools.

Above all, never under-estimate the power of visual analytics. A well thought out plot can save you paragraphs of explanation.

# References